---
title: "Multimodal Models and Affordances"
date: "September 12, 2024"
output:
  # pdf_document: 
  #    fig_caption: yes
  #    keep_md: yes
  #    keep_tex: yes
  html_document:
     keep_md: yes
     toc: yes
     toc_float: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```

```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(ggridges)
library(lmerTest)
library(broom.mixed)
library(ggrepel)
library(tools)
library(viridis)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```


# Load data

## Open HuggingFace models

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/llm_affordances/src/main_study/")
directory_path <- "../../data/processed/models/hf_models/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_hf_models <- bind_rows(csv_list)

df_hf_models = df_hf_models %>%
  mutate(condition = str_to_title(sub("_.*", "", condition)))

table(df_hf_models$model_name)
```

## Load closed-source models

```{r}
directory_path <- "../../data/processed/models/closed_models//"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_closed_models <- bind_rows(csv_list)

df_closed_models = df_closed_models %>%
  mutate(condition = str_to_title(sub("_.*", "", condition)))

table(df_closed_models$model_name)
```

## Merge

```{r}
df_merged = df_hf_models %>%
  bind_rows(df_closed_models)
```



# Confirmatory Analyses

```{r}

### Z-score response
df_hf_models = df_hf_models %>%
  group_by(model_name, version) %>%
  mutate(response_z = scale(response))

### Average z-scored response
df_summary <- df_merged %>%
  group_by(condition, model_name, num_params) %>%
  summarize(avg_response = mean(response_z, na.rm = TRUE),
            se_response = sd(response_z, na.rm = TRUE) / sqrt(n()))

### Plot avg z-scored response by condition
ggplot(df_summary, aes(x = model_name,
                       y = avg_response, 
                       color = condition, group = condition)) +
  geom_point(size = 3, 
             position = position_dodge(width = 0.5)) +  
  geom_errorbar(aes(ymin = avg_response - se_response, 
                    ymax = avg_response + se_response), 
                width = 0.2,
                position = position_dodge(width = 0.5)) + 
  labs(# title = "",
       x = "",
       y = "Z-scored Response",
       color = "") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal() +
  scale_color_viridis_d() +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "bottom")

ggplot(df_merged, aes(x = model_name, y = response_z, color = condition)) +
  # Plot raw data points with slight jitter for better visibility
  geom_jitter(size = 1, width = 0.2, height = 0, alpha = 0.1) +
  # Plot the means
  stat_summary(
    fun = mean,
    geom = "point",
    size = 3,
    position = position_dodge(width = 0.5)
  ) +
  # Plot the error bars (standard error)
  stat_summary(
    fun.data = mean_se,  # Calculate mean and standard error
    geom = "errorbar",
    width = 0.2,
    position = position_dodge(width = 0.5)
  ) +
  # Aesthetic adjustments
  labs(
    x = "",
    y = "Z-scored Response",
    color = ""
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal() +
  scale_color_viridis_d() +
  theme(
    axis.title = element_text(size = rel(1.2)),
    axis.text = element_text(size = rel(1.2)),
    legend.text = element_text(size = rel(1.2)),
    strip.text.x = element_text(size = rel(1.2)),
    legend.position = "bottom"
  )

```



## Afforded (non-canonical) vs. Non-Afforded 

```{r}
# Define a function to fit the lmer model and return the tidy summary
fit_and_tidy_lmer_model <- function(df) {
  # Fit the mixed model
  model_fit <- lmer(response_z ~ condition + prompt_type + version + (1 | group_id), 
                    data = filter(df, condition != "Related"), 
                    REML = FALSE)
  
  # Tidy the model (extracting coefficients, p-values, etc.)
  model_tidy <- tidy(model_fit)
  
  # Return both the model fit and the tidy dataframe
  list(model_fit = model_fit, model_tidy = model_tidy)
}

# Apply the function to each model_name group
model_results <- df_merged %>%
  group_by(model_name) %>%
  nest() %>%  # Nest data by model_name
  mutate(model_info = map(data, fit_and_tidy_lmer_model))  # Fit the model and get tidy output

# Extract the model fits and the tidy dataframes for each model
model_fits <- model_results %>%
  mutate(model_fit = map(model_info, "model_fit"),
         model_tidy = map(model_info, "model_tidy"))

# If you want to unnest the tidy summaries into one dataframe:
tidy_results <- model_fits %>%
  unnest(model_tidy)


tidy_results %>%
  filter(term == "conditionNon-Afforded") %>%
  ggplot(aes(x = reorder(model_name, estimate),
             y = estimate)) +
  geom_point(size = 3, alpha = .6) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_errorbar(aes(ymin = estimate - 2*std.error, 
                    ymax = estimate + 2*std.error), 
                alpha = .2,
                width=.2) + 
                # position=position_dodge(.9)) +
  labs(x = "Predictor",
       y = "Estimate",
       title = "Non-Afforded vs. Afforded") +
  theme_minimal() 

just_condition = tidy_results %>%
  filter(term == "conditionNon-Afforded") 

final_df = data.frame(
  model = just_condition$model_name,
  original_p = just_condition$p.value,
  adjusted_p_holm = p.adjust(just_condition$p.value, method = "holm")
)
final_df
```

## Related (canonical) vs. Non-Afforded 

```{r}
# Define a function to fit the lmer model and return the tidy summary
fit_and_tidy_lmer_model <- function(df) {
  # Fit the mixed model
  model_fit <- lmer(response_z ~ condition + prompt_type + 
                      version + (1 | group_id), 
                    data = filter(df, condition != "Afforded"), 
                    REML = FALSE)
  
  # Tidy the model (extracting coefficients, p-values, etc.)
  model_tidy <- tidy(model_fit)
  
  # Return both the model fit and the tidy dataframe
  list(model_fit = model_fit, model_tidy = model_tidy)
}

# Apply the function to each model_name group
model_results <- df_merged %>%
  group_by(model_name) %>%
  nest() %>%  # Nest data by model_name
  mutate(model_info = map(data, fit_and_tidy_lmer_model))  # Fit the model and get tidy output

# Extract the model fits and the tidy dataframes for each model
model_fits <- model_results %>%
  mutate(model_fit = map(model_info, "model_fit"),
         model_tidy = map(model_info, "model_tidy"))

# If you want to unnest the tidy summaries into one dataframe:
tidy_results <- model_fits %>%
  unnest(model_tidy)


tidy_results %>%
  filter(term == "conditionRelated") %>%
  ggplot(aes(x = reorder(model_name, estimate),
             y = estimate)) +
  geom_point(size = 3, alpha = .6) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_errorbar(aes(ymin = estimate - 2*std.error, 
                    ymax = estimate + 2*std.error), 
                alpha = .2,
                width=.2) + 
                # position=position_dodge(.9)) +
  labs(x = "Model",
       y = "Estimate",
       title = "Canonical vs. Non-Afforded") +
  theme_minimal()

just_condition = tidy_results %>%
  filter(term == "conditionRelated") 

final_df = data.frame(
  model = just_condition$model_name,
  original_p = just_condition$p.value,
  adjusted_p_holm = p.adjust(just_condition$p.value, method = "holm")
)
final_df
```


# Exploratory Analyses

# Accuracy

## Closed-source models

Although both GPTs show an effect *on average* between Afforded and Non-Afforded, they do assign a rating of "1" to many Afforded items.

```{r}
df_closed_models %>%
  ggplot(aes(x = response, fill = condition)) +
  geom_bar(stat = "count") +
  theme_minimal() +
  facet_wrap(~model_name + condition) +
  scale_fill_viridis_d() +
  labs(x = "Response") +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "none")
```


## All together

A slightly different distribution of models is found when you compare the *average difference* between Afforded vs. Non-Afforded as opposed to the *proportion of items in which Afforded > Non-AffordeD* (how "accuracy" is operationalized here).

```{r}
df_merged = df_merged %>%
  group_by(model_name, version) %>%
  mutate(response_z = scale(response))

df_accuracy = df_merged %>%
  select(condition, response_z, group_id, text, 
         model_name) %>%
  #filter(condition != "Related") %>%
  pivot_wider(
    names_from = condition,
    values_from = c(response_z),
    names_prefix = "condition_"
  ) %>%
  mutate(diff_main = condition_Afforded - `condition_Non-Afforded`,
         diff_manipulation_check = condition_Related -`condition_Non-Afforded`) %>%
  mutate(accuracy_main = diff_main > 0,
         accuracy_manipulation_check = diff_manipulation_check > 0)
  

df_accuracy %>%
  ggplot(aes(x = group_id, y = diff_main)) +
  geom_bar(stat = "identity") +
  # geom_vline(xintercept = 0, linetype = "dotted") +
  facet_wrap(~model_name) +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(x = "Item Number",
       y = "Difference (Afforded vs. Non-Afforded)") +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "none")


mean(df_accuracy$accuracy_main)
mean(df_accuracy$accuracy_manipulation_check)


summary(df_accuracy$diff_main)
sd(df_accuracy$diff_main)
summary(df_accuracy$diff_manipulation_check)
sd(df_accuracy$diff_manipulation_check)
  

accuracy_by_model = df_accuracy %>%
  group_by(model_name) %>%
  summarise(avg_accuracy_main = mean(accuracy_main),
            avg_accuracy_manipulation = mean(accuracy_manipulation_check))
accuracy_by_model


accuracy_by_model %>%
  ggplot(aes(x = reorder(model_name, avg_accuracy_main),
             y = avg_accuracy_main)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d() +
  labs(x = "Model",
       y = "Accuracy (Afforded vs. Non-Afforded)") +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "none")

accuracy_by_model %>%
  ggplot(aes(x = reorder(model_name, avg_accuracy_main),
             y = avg_accuracy_manipulation)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d() +
  labs(x = "Model",
       y = "Accuracy (Canonical vs. Non-Afforded)") +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "none")

```


## Cross-architecture analysis

```{r}
df_hf_models = df_hf_models %>%
  mutate(architecture = case_when(
    model_name %in% c("vilt-coco", "bridgetower", "vilt-f30k") ~ "Fusion",
    model_name %in% c("flava-full") ~ "Both",
    TRUE ~ "Dual-Encoder"
  ))

df_summary <- df_hf_models %>%
  group_by(condition, architecture) %>%
  summarize(avg_response = mean(response_z, na.rm = TRUE),
            se_response = sd(response_z, na.rm = TRUE) / sqrt(n()))

ggplot(df_summary, aes(x = architecture, y = avg_response, 
                       color = condition, group = condition)) +
  geom_point(size = 3, 
             position = position_dodge(width = 0.5)) +  
  geom_errorbar(aes(ymin = avg_response - se_response, 
                    ymax = avg_response + se_response), 
                width = 0.2,
                position = position_dodge(width = 0.5)) + 
  labs(# title = "",
       x = "Architecture",
       y = "Z-scored Response",
       color = "") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal() +
  scale_color_viridis_d() +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "bottom") 
```

## Scale Analysis

```{r}
df_hf_models = df_hf_models %>%
  mutate(architecture = case_when(
    model_name %in% c("vilt-coco", "bridgetower", "vilt-f30k") ~ "Fusion",
    model_name %in% c("flava-full") ~ "Both",
    TRUE ~ "Dual-Encoder"
  ))



df_summary <- df_hf_models %>%
  group_by(architecture, num_params, condition) %>%
  summarize(
    avg_response = mean(response_z, na.rm = TRUE),
    se_response = sd(response_z, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  ) %>%
  pivot_wider(
    names_from = condition,
    values_from = c(avg_response, se_response),
    names_prefix = "condition_"
  ) %>%
  mutate(
    avg_effect = avg_response_condition_Afforded - `avg_response_condition_Non-Afforded`,
    avg_manipulation_check = avg_response_condition_Related - `avg_response_condition_Non-Afforded`,
    se_effect = sqrt(se_response_condition_Afforded^2 + `se_response_condition_Non-Afforded`^2),
    se_manipulation_check = sqrt(se_response_condition_Related^2 + `se_response_condition_Non-Afforded`^2),
  )


ggplot(df_summary, aes(x = num_params, y = avg_effect, 
                       color = architecture)) +
  geom_point(size = 6, alpha = .5) +  
  geom_errorbar(aes(ymin = avg_effect - se_effect, 
                    ymax = avg_effect + se_effect), 
                width = 0.2) + 
  labs(# title = "",
       x = "Number of Parameters",
       y = "Effect Size",
       color = "") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal() +
  scale_color_viridis_d() +
  scale_y_continuous(limits = c(-1, 1)) +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "bottom") 

ggplot(df_summary, aes(x = num_params, y = avg_manipulation_check, 
                       color = architecture)) +
  geom_point(size = 6, alpha = .5) +  
  geom_errorbar(aes(ymin = avg_manipulation_check - se_manipulation_check, 
                    ymax = avg_manipulation_check + se_manipulation_check), 
                width = 0.2) + 
  labs(# title = "",
       x = "Number of Parameters",
       y = "Effect Size",
       color = "") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal() +
  scale_color_viridis_d() +
  scale_y_continuous(limits = c(-1, 1)) +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "bottom") 
```


## Role of `version` and `prompt_type`

```{r}
mod_full = lmer(data = filter(df_hf_models, condition != "Afforded"), 
                response_z ~ condition * version + condition * prompt_type+ 
                  (1 | group_id) + (1 | model_name), 
                    REML = FALSE)

df_summary <- df_hf_models %>%
  group_by(condition, model_name, prompt_type) %>%
  summarize(avg_response = mean(response_z, na.rm = TRUE),
            se_response = sd(response_z, na.rm = TRUE) / sqrt(n()))

ggplot(df_summary, aes(x = model_name, y = avg_response, 
                       color = condition, group = condition)) +
  geom_point(size = 3, 
             position = position_dodge(width = 0.5)) +  
  geom_errorbar(aes(ymin = avg_response - se_response, 
                    ymax = avg_response + se_response), 
                width = 0.2,
                position = position_dodge(width = 0.5)) + 
  labs(# title = "",
       x = "",
       y = "Z-scored Response",
       color = "") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal() +
  scale_color_viridis_d() +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "bottom") +
  facet_wrap(~prompt_type)

df_summary <- df_hf_models %>%
  group_by(condition, model_name, version) %>%
  summarize(avg_response = mean(response_z, na.rm = TRUE),
            se_response = sd(response_z, na.rm = TRUE) / sqrt(n()))

ggplot(df_summary, aes(x = model_name, y = avg_response, 
                       color = condition, group = condition)) +
  geom_point(size = 3, 
             position = position_dodge(width = 0.5)) +  
  geom_errorbar(aes(ymin = avg_response - se_response, 
                    ymax = avg_response + se_response), 
                width = 0.2,
                position = position_dodge(width = 0.5)) + 
  labs(# title = "",
       x = "",
       y = "Z-scored Response",
       color = "") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal() +
  scale_color_viridis_d() +
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        # legend.title = element_text(size = rel(1.5)),
        strip.text.x = element_text(size = rel(1.2)),
        legend.position = "bottom") +
  facet_wrap(~version)

```

